Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-08 VegOzone.tex
Change doys' position

\correspondence{Stefanie Falk (stefanie.falk@geo.uio.no)}

@@

%Ozone
The derived ozone climatology (\chem{\left<[O_3]\right>}) (1986--1996) is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data to guide the eye. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display the expected, pronounced seasonal cycle. The spring peak occurs on average on $100\,\unit{doy}$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. The decline in \chem{\left<[O_3]\right>} coincides with the average begin of \chem{CO_2} uptake by coniferous trees \citep{TB:Kolari2007, TP:Wallin2013}. In July--September, ozone is occasionally almost completely depleted. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

@@

Another important factor is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product \citep{MODIS_PSN} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is, hence, one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR). For comparison, the generic values for central Europe ($A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

\begin{table}[t]
  \caption{Start and end of growing season in units of \unit{doy}.}
  \label{tab:sensitivity_tests_gs}
  \begin{tabular}{lcll}
    \tophline
    Species & Year & $A_\mathrm{start}$ & $A_\mathrm{end}$\\
    \middlehline
    Default$^b$ & -- & 100 & 307\\
    \multirow{2}{*}{Deciduous tree} & 2018 & 129$^*$ & 261$^a$ \\
    & 2019 & 130$^c$ & 274$^a$ \\
    \multirow{2}{*}{Coniferous tree} & 2018 & 122$^a$ & 261$^a$ \\
    & 2019 & 106$^a$ & 274$^a$ \\
    \multirow{2}{*}{Perennial grassland} & 2018 & 159$^d$ & 261$^a$\\
    & 2019 & 161$^d$ & 274$^a$ \\
    \bottomhline
  \end{tabular}
  \belowtable{$^a$ MODIS (Aqua/Terra) net photosynthesis product;\\
    $^b$ \citet{ACP:Simpson2012},\\
    $^c$ $5\,\unit{days}-5\,\unit{^\circ C}$-rule, $T_\mathrm{air}$ from \href{seNorge.no}{seNorge.no};\\
    $^d$ One month after snow melt; reference station Øvre Neiden (Sør-Varanger, NOR)} % Table Footnotes
\end{table}

@@

We have studied the effect of bespoke parameterizations of common vegetation types (coniferous and deciduous trees, perennial grassland) to a subarctic climate on $\mathrm{POD_y}$. The comparison between meteorological conditions in 2018 and 2019 and their divergence from climatology allowed us to assess the influence of key environmental variables such as temperature, PPFD, and precipitation on vegetation susceptibility to \chem{O_3} damage in light of future changes as may occur under climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden. However, ozone concentrations during the growing season were not significantly enhanced and \chem{POD_1} values were very similar for both years.

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The deduced start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. We determined the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one month latency period after snow melt which is supported by observational data from Finland \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season which is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{figB1}
  \includegraphics[width=12cm]{DO3SE_results_pody_gsto_o3.png}
  \caption{$\mathrm{DO_3SE}$ modeling results for mapping manual default parameterization. $\mathrm{POD_y}$ is shown over day of year (\unit{doy}), March--October. A flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) has been chosen. \chem{[O_3]} are plotted on the same axis and scales as $G_\text{sto}^\text{leaf}$ but in units of $\unit{ppb}$. (a, b) deciduous tree; (c, d) coniferous tree; (e, f) perennial grassland. From left to right: 2018, 2019.}
  \label{fig:pody_mm_composit}
\end{figure*}

@@

\begin{figure}[th]
  %\includegraphics[width=8.3cm]{figB2}
  \includegraphics[width=8.3cm]{modis_Psn.pdf}
  \caption{Estimated $G_\mathrm{start}$ and $G_\mathrm{end}$ of growing season for coniferous trees from MODIS Aqua/Terra net photosynthesis (PSN) product. A $1\times 1\,\unit{km}$ area around Svanhovd was selected. Daily averaged data for both 2018 and 2019 has been fitted with a quadratic polynomial function. The numerically computed root yields: BGS 122/106 day of year (\unit{doy}) and EGS 261/274 day of year for 2018/2019, respectively.}
  \label{fig:modis_Psn}
\end{figure}

>>>>

\correspondence{Stefanie Falk (stefanie.falk@lmu.de)}

@@

%Ozone
The derived ozone climatology (\chem{\left<[O_3]\right>}) (1986--1996) is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data to guide the eye. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display the expected, pronounced seasonal cycle. The spring peak occurs on average on day of the year (\unit{doy}) $100$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. The decline in \chem{\left<[O_3]\right>} coincides with the average begin of \chem{CO_2} uptake by coniferous trees \citep{TB:Kolari2007, TP:Wallin2013}. In July--September, ozone is occasionally almost completely depleted. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

@@

Another important factor is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product \citep{MODIS_PSN} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start}$ as \unit{doy}~122 for 2018 and \unit{doy}~106 in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is, hence, one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR). For comparison, the generic dates for central Europe ($A_\mathrm{start} = 100$ and $A_\mathrm{end} = 307$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

\begin{table}[t]
  \caption{Start and end of growing season in \unit{doy}.}
  \label{tab:sensitivity_tests_gs}
  \begin{tabular}{lcll}
    \tophline
    Species & Year & $A_\mathrm{start}$ & $A_\mathrm{end}$\\
    \middlehline
    Default$^b$ & -- & 100 & 307\\
    \multirow{2}{*}{Deciduous tree} & 2018 & 129$^*$ & 261$^a$ \\
    & 2019 & 130$^c$ & 274$^a$ \\
    \multirow{2}{*}{Coniferous tree} & 2018 & 122$^a$ & 261$^a$ \\
    & 2019 & 106$^a$ & 274$^a$ \\
    \multirow{2}{*}{Perennial grassland} & 2018 & 159$^d$ & 261$^a$\\
    & 2019 & 161$^d$ & 274$^a$ \\
    \bottomhline
  \end{tabular}
  \belowtable{$^a$ MODIS (Aqua/Terra) net photosynthesis product;\\
    $^b$ \citet{ACP:Simpson2012},\\
    $^c$ $5\,\unit{days}-5\,\unit{^\circ C}$-rule, $T_\mathrm{air}$ from \href{seNorge.no}{seNorge.no};\\
    $^d$ One month after snow melt; reference station Øvre Neiden (Sør-Varanger, NOR)} % Table Footnotes
\end{table}

@@

To summarize, we have studied the effect of bespoke parameterizations of common vegetation types (coniferous and deciduous trees, perennial grassland) to a subarctic climate on $\mathrm{POD_y}$. The comparison between meteorological conditions in 2018 and 2019 and their divergence from climatology allowed us to assess the influence of key environmental variables such as temperature, PPFD, and precipitation on vegetation susceptibility to \chem{O_3} damage in light of future changes as may occur under climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden. However, ozone concentrations during the growing season were not significantly enhanced and \chem{POD_1} values were very similar for both years.

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We deduced the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one month latency period after snow melt which is supported by observational data from Finland \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season which is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{figB1}
  \includegraphics[width=12cm]{DO3SE_results_pody_gsto_o3.png}
  \caption{$\mathrm{DO_3SE}$ modeling results for mapping manual default parameterization. $\mathrm{POD_y}$ is shown over \unit{doy}, March--October. A flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) has been chosen. \chem{[O_3]} are plotted on the same axis and scales as $G_\text{sto}^\text{leaf}$ but in units of $\unit{ppb}$. (a, b) deciduous tree; (c, d) coniferous tree; (e, f) perennial grassland. From left to right: 2018, 2019.}
  \label{fig:pody_mm_composit}
\end{figure*}

@@

\begin{figure}[th]
  %\includegraphics[width=8.3cm]{figB2}
  \includegraphics[width=8.3cm]{modis_Psn.pdf}
  \caption{Estimated $G_\mathrm{start}$ and $G_\mathrm{end}$ of growing season for coniferous trees from MODIS Aqua/Terra net photosynthesis (PSN) product. A $1\times 1\,\unit{km}$ area around Svanhovd was selected. Daily averaged data for both 2018 and 2019 has been fitted with a quadratic polynomial function. The numerically computed root yields: BGS \unit{doy}~122/106 and EGS \unit{doy}~261/274 for 2018/2019, respectively.}
  \label{fig:modis_Psn}
\end{figure}
